---
title: "R_Week_5_Hmw_607_Tidy_Transform_Data_Daniel_Thonn"
output: html_document
course: R-607 CUNY MSDA program

---

## Assignment 607_Homework-5: Tidy and Transform data

Summary of Assignment
This assignment involves Tidying and Transforming flight data

This Assignment requires the following:

1). R-Studio

The following R-packages are used:
1.tidyr
2.dplyr
3.ggplot2 

Steps to reproduce:
1). run the R-Studio file: R_607_Week_5a_Hmk_Tidy_Transform_Data_Daniel_Thonn.Rmd


Setting up and Preparing the Environment

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE}

#install.packages("stringr")
#install.packages("tidyr")
#install.packages("dplyr")
#install.packages("plyr")
#library(stringr)
library(tidyr)
library(dplyr)
#library(plyr)
#detach("package:plyr", unload=TRUE)
library(ggplot2)

```


Load the character names data into an R dataframe, cleanup, and convert to long format


```{r, echo=TRUE}
#Load data from a text file
#data_1 <- readLines("C:/mysqldata/tournamentinfo.txt") 
#head(data_1)

# Load the data csv file to a vector
a <- read.csv(
              "C:/mysqldata/Arrivals_Airlines.csv",
              sep=",",
              na.strings = "",
              blank.lines.skip = TRUE,
                stringsAsFactors=FALSE)
list(a)

# Convert the input vector to a dataframe
df_arrivals_1 = data.frame(a)
list(df_arrivals_1)

# Remove the empty row [3]
df_arrivals_2 <- df_arrivals_1[-3,]
list(df_arrivals_2)
names(df_arrivals_2)

# Label columns for Airlines and Status
df_arrivals_2b <- df_arrivals_2
names(df_arrivals_2b)[names(df_arrivals_2b) == "X"] <- "Airlines"
names(df_arrivals_2b)[names(df_arrivals_2b) == "X.1"] <- "Status"
list(df_arrivals_2b)

# Remove NA's
df_arrivals_3 <- replace(df_arrivals_2b, is.na(df_arrivals_2), "") 
list(df_arrivals_3)

df_arrivals_4 <- df_arrivals_3
names(df_arrivals_4)
list(df_arrivals_4)

length(df_arrivals_4$Airlines)

df_arrivals_5 <- df_arrivals_4
list(df_arrivals_5)

# Fill in missing column names
for (i in 1:length(df_arrivals_5$Airlines))
{
  if (i %% 2){
    next
  }
  #print(df_arrivals_4$Airlines[i])
  df_arrivals_5$Airlines[i] <- paste(df_arrivals_4$Airlines[i],df_arrivals_4$Airlines[i-1])
}

list(df_arrivals_5)

# Remove spaces from column Airlines values
df_arrivals_5$Airlines <- gsub(" ", "", df_arrivals_5$Airlines)
df_arrivals_5$Airlines

list(df_arrivals_5)

# Convert wide dataframe into long dataframe
df_arrivals_long <- gather(df_arrivals_5, City, NumFlights, Los.Angeles:Seattle) %>% mutate(City = gsub( "\\.", " ", City)) %>% arrange(Airlines)

#list(df_arrivals_long)


```



Use dplyer to analyze the flight numbers and plot the results

```{r, echo=TRUE}

list(df_arrivals_long)

# Add a column for percent of flights delayed or ont-time relative to total number of flights
By_Airlines_df_0a <- mutate(df_arrivals_long,percent1= round(NumFlights/sum(NumFlights) * 100, 1))
By_Airlines_df_0a

# Setup group by criteria by Airlines
By_Airlines_df_1 <- group_by(By_Airlines_df_0a,Airlines,Status) 
By_Airlines_df_1

# Select columns for analysis dataframe by Airlines
By_Airlines_df_1a <- select(By_Airlines_df_1,Airlines,Status,NumFlights,percent1)
By_Airlines_df_1a

# Present columns based on select and group-by criteria by Airlines
By_Airlines_df_3 <- summarise(By_Airlines_df_1a,SumNumFlights = sum(NumFlights),SumPercent = paste0(sum(percent1),"%"))
By_Airlines_df_3

# Graph delayed flights per total flights by Airline
By_Airlines_df_3 %>% filter(Status == "delayed") %>%
    ggplot( aes(x=Airlines, y=SumPercent, fill=Airlines)) +
    geom_bar(stat="identity", position=position_dodge(), colour="black", width = 0.5) +
    coord_flip() + 
    ggtitle("Percent of Delays per Total Flights by Airline") +
    xlab("Airline") + ylab("Percentage of Flights Delayed") +
    geom_text(aes(label=paste(SumPercent)), vjust=0.5, hjust=1.1,color="black")


# Graph on-time flights per total flights by Airline
By_Airlines_df_3 %>% filter(Status == "on time") %>%
    ggplot( aes(x=Airlines, y=SumPercent, fill=Airlines)) +
    geom_bar(stat="identity", position=position_dodge(), colour="black", width = 0.5) +
    coord_flip() + 
    ggtitle("Percent of On-Time Trips per Total Flights by Airline") +
    xlab("Airline") + ylab("Percentage of Flights On-Time") +
    geom_text(aes(label=paste(SumPercent)), vjust=0.5, hjust=1.1,color="black")

# Setup group by criteria by Airlines and City
By_Airlines_df_1b <- group_by(By_Airlines_df_0a,Airlines,Status,City) 
By_Airlines_df_1b

# Select columns for analysis dataframe by Airlines and City
By_Airlines_df_1c <- select(By_Airlines_df_1b,Airlines,Status,City,NumFlights,percent1)
By_Airlines_df_1c

# Present columns based on select and group-by criteria by Airlines
By_Airlines_df_3b <- summarise(By_Airlines_df_1c,SumNumFlights = sum(NumFlights),SumPercent = paste0(sum(percent1),"%"))
By_Airlines_df_3b

# Graph delayed flights per total flights by Airline and City
By_Airlines_df_3b %>% filter(Status == "delayed") %>%
    ggplot( aes(x=City, y=SumPercent, fill=Airlines)) +
    geom_bar(stat="identity", position=position_dodge(), colour="black", width = 0.5) +
    coord_flip() + 
    ggtitle("Percent of Delays per Total Flight by Airline by City") +
    xlab("City") + ylab("Percentage of Flights Delayed") +
    geom_text(aes(label=paste(SumPercent)), vjust=0.5, hjust=1.1,color="black")


# Graph on-time flights per total flights by Airline
By_Airlines_df_3b %>% filter(Status == "on time") %>%
    ggplot( aes(x=City, y=SumPercent, fill=Airlines)) +
    geom_bar(stat="identity", position=position_dodge(), colour="black", width = 0.5) +
    coord_flip() + 
    ggtitle("Percent of On-Time Trips per Total Flights by Airline by City") +
    xlab("City") + ylab("Percentage of Flights On-Time") +
    geom_text(aes(label=paste(SumPercent)), vjust=0.5, hjust=1.1,color="black")


```

Conclusion: Overall AMWEST has the higher total proportion of delayed flights at 7.3% versus 4.5% for ALASKA.  But AMWEST also had the higher number of on-time flights at 58.5% versus 29.7% for ALASKA.  The number of delays versus on-time was 7.3/58.5 = 12.4% for AMWEST and 4.5/29.7 = 15.1% for ALASKA, so ALASKA has a higher relative proportion of delayed flights. 

In analysis by City Seattle has the highest total proportion of delayed flights for ALASKA airlines (2.8%), and Phoenix the highest number of delays by AMWEST (3.8%).  

By City Pheonix also has the highest total proportion of on-time flights at a total of 46% for both Airlines combined.

If we compare delays to on-time by city the breakdown of relative proportional delays is as follows:

Los Angeles: 1.7/4.8 = 35.4%
Seattle:  3.4/11.5 = 29.6%
San Francisco:  2.1/10.9 = 19.2%
San Diego:  .7/5.4 = 12.9%
Phoenix: 3.9/46 = 8.4%

This shows the breakdown of relative delays versus on-time flights by city and shows Los Angeles is the highest proportion and Phoenix has the least delays for both airlines. 












#END
